A weakly structured stem for human origins in Africa

### 2022-02-25 paper.tex
IICR rewrite

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events. 
Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). These constraints allowed us to integrate information from previous \sgcomment{genetic and archaelogical?} research,
to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).

@@

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above \sgcomment{XX} 
        *: The ancestors of the Nama diverged from other African groups
        $\sim$120--135kya, which later branch into West and East African
        groups 60kya. Migration rates and durations are shown for 1) East Africans
        and their ancestors, and 2) the Mende, respectively.
    }
    \label{tab:migration-rates}
\end{table}

@@

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related, \sgcomment{and each branch contributes large amounts to all contemporary populations} (Figure 3). 
Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
to this early population structure (Figure 3 and supplementary text)

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time, which are commonly interpreted as effective
population size histories. Trajectories of $N_e$ from prior studies strongly
support a bottleneck in Out of Africa populations. African populations also dip
in effective size around 100-200kya, although this decrease is more modest than
the OOA event \citep{Mallick2016-lx}. One possible explanation is that
ancestral population structure during the Middle Stone Age inflated $N_e$ and
subsequent gene flow among ancestral Africans resulted in a more recent
decreased $N_e$ \citep{Mazet2016-wn}.
%To our knowledge, explicit comparisons of this hypothesis with SMC-style
%approaches have not been published.
We simulated genomic data under four inferred demographic models to compare
with observed coalescence rate histories. All models, including the
single-origin model, generate multi-modal curves indicating that the ancestral
increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
rate in humans, although none of the models perfectly match the exact timing of
the early fluctuations in $N_e$ observed in the data.
%Simulated Relate Ne curves under the CM and MM are distinguished by having 2
%and 3 modes, respectively.  The Continuous Migration model had the better
%qualitative fit with the data, but the ancestral mode was shifted further back
%in time relative to the data.
Each simulated model also underestimated the level of population growth during
the Holocene, indicating additional free parameters are necessary to refine
these growth rates. \sgcomment{I am not sure I am happy with these interpretations 
-- let's discuss. I feel like this paragraph does not support our data well, and a likely culprit 
in explaining the differences is the inaccuracy of TMRCA estimates from real data.}

@@

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly-structured-stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text \sgcomment{X})\aprcomment{Will we still include this?} \sgcomment{yes!}, even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Identifying the morphological signature of ancient
structure in the fossil record will thus require careful analysis of many broadly
similar remains. Morphologically distinct
crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
contributed appreciably to modern human ancestries.

>>>>

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events. 
Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). These constraints allowed us to integrate information from previous genetic and archaelogical research,
to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).

@@

\begin{table}[t!]
    \centering
    \scriptsize
    \begin{tabular}{lllrrrr}
        \toprule
        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
        & & & 
            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
        \midrule
        \textbf{Continuous Model} & & & & & \\
        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
        \textbf{Merger Model} & & & & & \\
        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
        \bottomrule
    \end{tabular}
    \caption{
        \textbf{Migration and divergence parameters from best fit models.}
        Labeled migration rates correspond to symmetric continuous migration
        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
        merger models inferred a relatively deep split of human stem branches,
        though they were connected by ongoing migration that maintained their
        genetic similarity. Bold text indicates migration rates above $1e-4$ 
        *: The ancestors of the Nama diverged from other African groups
        $\sim$120--135kya, which later branch into West and East African
        groups 60kya. Migration rates and durations are shown for 1) East Africans
        and their ancestors, and 2) the Mende, respectively.
    }
    \label{tab:migration-rates}
\end{table}

@@

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
Because of this 
relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
to this early population structure (Figure 3 and supplementary text)

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred models. Many 
previous studies have found a reduction of coalescence rates 
between $1000$kya and $100$kya in humans, and thus inferred an increase 
in $N_e$ during the same period. \citep{Li, Durbin}. 
This increase in $N_e$ could be due to either an increase in population or to 
ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
All models, including the
single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim} and supplementary section \sgcomment{X}). Whereas the single-origin model achieves this by
an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.

@@

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly-structured-stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text \sgcomment{X}), even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Identifying the morphological signature of ancient
structure in the fossil record will thus require careful analysis of many broadly
similar remains. Morphologically distinct
crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
contributed appreciably to modern human ancestries.
